# Violation of CLRM Assumptions

## The CLRM Assumptions Again

We have seen the assumptions for the Ordinary Least Squares (OLS) estimators $\hat{\alpha}$ and $\hat{\beta}'s$ to have desirable properties. What does it mean that assumptions do not hold?.

Consider the following bivariate population model:

$$Y_i = \alpha + \beta_1 X_{1i} + \beta_2 X_{2i} + u_i$$

**Violation of Assumption 4: Endogeneity**

We start with this one, as it is the most commons and serious of many econometric specifications. Basically we have endogeneity when any of the explanatory variable $X's$ is correlated with the error term $u$.

$$Cov(X_i's, u_i) \neq 0$$

-   This basically means that there are other important factors that affect the dependent variable $Y_i$ which if not included in the regression model will result in incorrect estimates!

-   **Implication:** Essentially $\hat{\beta}$ will be biased and inconsistent.

-   **Violation of Assumption 1: Conditional mean is not zero**

-   Endogeniety or violation of assumption 4 implies that Assumption 1, the zero conditional mean, also does not hold! The error term $u$ will not have expected value of zero, **given the values of the explanatory variables** $X's$.

$$E(u_i | X_i's) \neq 0$$

**Violation of Assumption 2: Heteroskedasticity**

The error term $u$ has non-constant variances given the values of the explanatory variables.

$$Var(u_i | X_i's) \neq \sigma^2$$

-   **Implication:** In this case, OLS estimates are still unbiased but their standard errors are incorrect/inefficient, which leads to faulty hypothesis tests and confidence intervals.

**Violation of Assumption 3: Autocorrelation**

The error terms for any two or more different observations are correlated.

$$Cov(u_i, u_j | X_i) \neq 0 \quad \text{for all } i \neq j$$

-   **Implication:** Like heteroskedasticity, if we have autocorrelation then OLS estimates remain unbiased but the standard errors are incorrect/inefficient, leading to unreliable inference.
-   **Note:** Autocorrelation (a.k.a. serial autocorrelation) is usually a concern for **time-series** data, i.e. usually $Cov(u_t , u\_{t-1}) \neq 0$.

**Violation of Assumption 5: Perfect (or near perfect) Multicollinearity**

-   **Interpretation (for simple regression):** In the simple regression model with one explanatory variable, this assumption is automatically satisfied as long as $X$ is not constant.
-   **Implication:** When we have perfect multicollinearity, any independent variable is a perfect linear combination of other independent variable(s), then we canot get OLS estimates, and econometric packages will often delete one.

## Checking Assumptions

### Testing for Heteroskedasticity

As previously mentioned, a quick look at residual plots is useful a good start to vidually look for hoteroskedasticity and/or non-zero mean.

```{r}
# Fit the model
model <- lm(mpg ~ wt + hp, data = mtcars)

# Create a dataframe of fitted values and residuals
diagnostic_data <- data.frame(
  fitted = fitted(model),
  residuals = resid(model)
)

# Plot residuals vs. fitted values
library(ggplot2)
ggplot(diagnostic_data, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs. Fitted Values",
       x = "Fitted Values (Y_hat)",
       y = "Residuals (u_hat)") +
  theme_minimal()
```

A more formal test is the **White test**, a general and powerful test for heteroskedasticity. The test works by regressing the squared residuals from the original model on the original explanatory variables, their squares, and their cross-products.

The null and alternative hypotheses are:

$H_0$: Homoskedasticity exists (the error variance is constant).

$H_1$: Heteroskedasticity exists (the error variance is not constant).

```{r}
#| warning: false
#| message: false

# install.packages("lmtest") # Uncomment and run if needed
library(lmtest)

# Perform the White test for the simple model mpg ~ wt
bptest(model, ~ wt * hp + I(wt^2) + I(hp^2), data = mtcars)
```

The `bptest()` output shows a BP test statistic and a p-value. A p-value \< 0.05 means you reject the null hypothesis of homoskedasticity. In the `mtcars` example above, we fail to reject the null hypothesis, suggesting that we have may not have heteroskedasticty.

Alternatively:

```{r}
#| warning: false
#| message: false

# 1. Estimate the original simple regression model
simple_model <- lm(mpg ~ wt, data = mtcars)

# 2. Obtain the squared residuals from the model
squared_residuals <- resid(simple_model)^2

# 3. Perform the "auxiliary regression" for the White test:
# Regress the squared residuals on the original regressor (wt) and its square (wt²).
white_aux_model <- lm(squared_residuals ~ wt + I(wt^2), data = mtcars)

# 4. Conduct an F-test on the auxiliary model.
# The null hypothesis is that the coefficients on 'wt' and 'I(wt^2)' are zero.
# install.packages("car") # Uncomment and run if you don't have the 'car' package
library(car)

linearHypothesis(white_aux_model, c("wt=0", "I(wt^2)=0"))
```

We look at the F-statistic and its p-value (Pr(\>F)). A p-value \< 0.05 provides evidence to reject the null hypothesis ($H_0$) of homoskedasticity, thereby suggesting the error variance is not constant and depends on weight (wt). Here, again Pr(\>F) is 0.5312, hence we fail to reject the null.

### Testing for Autocorrelation


### Testing for Multicollinearity


### Testing for Endogeneity

Endogeneity occurs when an explanatory variable is correlated with the error term ($Cov(X, u) \neq 0$), violating a key Gauss-Markov assumption (4). This often arises from:

1.  Confounding leading to omitted variable bias
2.  Reverse causality
3.  Measurement Error
4.  Simultaneity

The consequence is that the OLS estimator becomes **biased and inconsistent**.

**The Logic of the Hausman Test**

The test follows a straightforward logic:

1.  **Null Hypothesis (**$H_0$): The variable in question is *exogenous* ($Cov(X, u) = 0$). OLS is consistent and efficient.

2.  **Alternative Hypothesis (**$H_1$): The variable is *endogenous* ($Cov(X, u) \neq 0$). OLS is inconsistent.

The test compares two estimators:

-   **OLS Estimator:** Efficient (has the smallest possible variance) under $H_0$, but **inconsistent** under $H_1$.
-   **IV (Instrumental Variables) Estimator:** Consistent under both $H_0$ and $H_1$, but **inefficient** (has larger variance) under $H_0$.

If the variable is exogenous ($H_0$ is true), the OLS and IV estimates should be similar. If they are significantly different, we have evidence that endogeneity is present ($H_1$ is true).

**Implementing the Hausman Test in R: A Step-by-Step Guide**

The test is implemented as a **Durbin-Wu-Hausman** test via a convenient auxiliary regression.

**Prerequisite:** You must have at least one **valid instrument** for the potentially endogenous variable. A valid instrument must be:

1.  **Relevant:** Correlated with the endogenous variable.
2.  **Exogenous:** *Not* correlated with the error term ($Cov(Z, u) = 0$).

**Scenario:** Suppose we fear that `wt` (weight) is endogenous in our model `mpg ~ wt`. Suppose we use `hp` (horsepower) as instruments.

**Step 1: Estimate the First Stage Regression**

Regress the potentially endogenous variable (`hp`) on all exogenous variables and the instruments.

```{r}
#| warning: false
#| message: false

# First Stage: Regress the endogenous variable on instruments and other exogenous vars
first_stage <- lm(hp ~ wt + hp, data = mtcars)

# Retrieve the residuals from the first stage
first_stage_residuals <- resid(first_stage)

# Add these residuals to the original dataset for the next step
mtcars$fs_resid <- first_stage_residuals
```

**Step 2: Estimate the Auxiliary Regression** Run the original model, but *include the first-stage residuals* as an additional regressor.

```{r}
#| warning: false
#| message: false

# Auxiliary Regression: Original model + first stage residuals
auxiliary_model <- lm(mpg ~ wt + fs_resid, data = mtcars)
summary(auxiliary_model)
```

**Step 3: Interpret the Result** - The key is the **t-test on the coefficient of the residual variable (`fs_resid`)**. - **Null Hypothesis (**$H_0$): The coefficient on the residuals is zero. This means the variable (`wt`) is exogenous. - **Alternative Hypothesis (**$H_1$): The coefficient on the residuals is *not* zero. This is evidence of endogeneity.

A low p-value (typically \< 0.05) on the `fs_resid` coefficient leads to a rejection of the null hypothesis, suggesting that `wt` is indeed endogenous.

**Using the `ivreg` and `lmtest` packages**

A more efficient method is to use the `ivreg()` function from the `AER` package and then formally test for endogeneity.

Specifying a model with instruments: The syntax y \~ x1 + x2 \| z1 + z2 means that we are regressing y on x1 and x2, using for x2 instruments z1 and z2 (and x1 is included as its own instrument).

```{r}
#| warning: false
#| message: false

# install.packages("AER") # Install the Applied Econometrics with R package
library(AER)
library(lmtest)

# 1. Estimate the model via IV and OLS
# IV model: Specify the formula and instruments
iv_model <- ivreg(mpg ~ wt | hp, data = mtcars)

# 2. Perform the Hausman test
# The null is that OLS is consistent (no endogeneity)

# Run summary with diagnostics
summary(iv_model, diagnostics = TRUE)
```

**Interpreting the Test:**

**Weak Instruments Test**

What it tests: Whether the instrument(s) (hp) is sufficiently correlated with the endogenous regressor (wt). A strong first-stage relationship is crucial for IV to work.

Interpretation: The null hypothesis is that the instruments are weak.

The result (p-value = 4.15e-05), an extremely low p-value, suggests that we should strongly reject the null. The instrument(s) are not weak; they are strong and relevant. This is a good sign.

**Wu-Hausman Test (for Endogeneity)**

What it tests: This is the test for endogeneity. The null hypothesis ($H_0$) is that the variable (wt) is exogenous. The alternative ($H_1$) is that it is endogenous.

Interpretation: A low p-value suggests you should reject $H_0$ and use the IV estimator. A high p-value means you cannot reject $H_0$ and should prefer the efficient OLS estimator.

The result (p-value = 0.0014), a low p-value means we should reject the null hypothesis. There is statistical evidence that wt is endogenous. Therefore, the standard OLS estimates for mpg \~ wt gives biased etimates.

**Sargan Test (for Overidentifying Restrictions)**

What it tests: This test checks the validity of your overidentifying instruments. It is only relevant if you have more instruments than endogenous variables. The null hypothesis ($H_0$) is that the extra instruments are valid (uncorrelated with the error term).

Interpretation: A low p-value is undesirable, as it means you should reject $H_0$ and suspect that at least one of your extra instruments is invalid.

**What to Do If You Find Endogeneity?**

If the test suggests endogeneity, you should **not trust the OLS results**. You must use a method that addresses the endogeneity, such as:

-   **Finding Better Controls:** If the endogeneity is from omitted variable bias.
-   **Instrumental Variables (IV) Regression:** The primary solution, implemented with `ivreg()`.
-   **Using Panel Data Methods:** Such as fixed effects models, if you have panel data.

**Important:** The validity of the Hausman test hinges on the **quality of your instruments**. If your instruments are weak or invalid, the test itself is unreliable. Always check the first-stage F-statistic to ensure instrument strength (a rule of thumb is F-stat \> 10).
